Quantifying Stability and Change in Personal Culture Using Panel Data

### 2023-09-24 manuscript.Rmd

A recurring focus in the sociology of culture is the question of whether people, as they move through their lives, change their personal culture –- the declarative and non-declarative attitudes, beliefs, values, practices, and dispositions [@lizardo2017; @kiley2020; @lersch2023].

>>>>

A recurring focus in the sociology of culture is the question of whether people, as they move through their lives and transition between social roles, change their personal culture –- the declarative and non-declarative attitudes, beliefs, values, practices, and dispositions that exist at the individual level [@lizardo2017; @kiley2020; @lersch2023].

### 2023-09-24 manuscript.tex

---
title: "Quantifying Stability and Change in Personal Culture Using Panel Data"
author:
- 
output:
  pdf_document:
    keep_tex: true
bibliography: kileybib.bib
csl: asa.csl
abstract: "Recent work has produced conflicting interpretations on the question of whether adults undergo durable changes in their personal culture over time. This paper asserts that this divergence is less due to researchers' findings than the approach taken -- a tournament of models focusing on the presence or absence of change. To move beyond this, we reanalyze the data used previously in this debate with an approach that quantifies the processes of cultural difference. In doing so, we compare how much cultural difference is explained by intrapersonal change and interpersonal differences. Our results harmonize recent findings by showing that, while most measures of personal culture show change over time, the variance explained by intrapersonal change is typically small compared to interpersonal differences at baseline. Quantifying these processes offers new ways to address theoretical questions of when and how their relative importance shifts. We examplify this with the case of college experience."
fontsize: 11pt
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F, message = F, warning = F, fig.align = "center")
```

# Introduction

A recurring focus in the sociology of culture is the question of whether people, as they move through their lives and transition between social roles, change their personal culture –- the declarative and non-declarative attitudes, beliefs, values, practices, and dispositions that exist at the individual level [@lizardo2017; @kiley2020; @lersch2023].

This seemingly simple question underlies a number of central debates across sociological subfields. For example, pragmatist theories of action claim that changes in social environments cause people to adapt their views and make new meanings [@swidler2001; @gross2009], while Bourdieusian practice theories argue that the “past conditions of production” leave a mark on people's personal culture that lasts throughout their lives [@bourdieu1990]. Similarly, models of social influence assume that people adapt their culture in the face of new information [@vaisey2010a], while the emphasis on cohort effects in models of aggregate social and cultural change requires them to be open to change while young but become fairly resistant to it as they age [@ryder1965]. Finally, life course theories posit slower but still important changes over time as people advance through important transitions in their lives [@elder2016; @bardi2011].

Because all of the above theoretical perspectives have some empirical support, the current theoretical debate revolves around not whether they exist, but when and how they each apply and the relative contribution of each to explaining cultural differences. That being said, scholars have struggled to reach a consensus on the relative importance of change during adulthood. Over the span of a few years, most apparent changes in people’s responses to personal culture items appear to be transitory, with little evidence of persistent change among adults [@kiley2020; @vaisey2021]. These results suggest change is not a strong force in explaining personal cultural differences. When looked at over longer time horizons, however, there is at least some evidence that adults make persistent cultural changes, especially as they move through distinct social roles across the life course [@lersch2023].

These seemingly inconsistent findings are partly due to the fact that researchers have taken a “tournament of models” approach in adjudicating the different perspectives [@lersch2023: p. 228]. Accordingly, researchers have formalized models of responses to cultural items over time as demonstrating either “change” or “no change” to then compare which model fits observed data better, thus functionally asking whether we ever observe within-person change in personal culture.

We argue, however, that asking only *whether* people change obscures what are actually highly compatible empirical patterns. Whether researchers are able to detect "change" in a given dataset depends not only on sample populations, item measurement, statistical methods and power, but crucially also on definitions of what counts as “change.” These methodological differences therefore limit the possibility of convergence on the theoretical debates. Moreover, asking whether these is evidence of change limits the kind of questions that can be asked to, “Do we see change in this measure?” But such inquiries are ill-suited to address what’s at the core of the theoretical debate. There is enough evidence to assume that neither socialization at an early age nor the influence of the contemporary environment alone can explain cultural differences. The question that remains is under what conditions, across which groups, and in what domains do we observe differences in the relative importance of intrapersonal change and interpersonal differences. This focus is necessarily comparative and requires an approach that quantifies forms of cultural change rather than calling victory for the one or the other perspective.

In this paper, we try to move beyond previous debates by moving away from the “tournament of models” approach toward a quantification of the processes that lead to differences in personal culture. Drawing on the seven panel datasets analyzed in these debates, we use established methods to quantify the amount of variance in responses that is due to between-people differences at a single point during the survey versus the amount of variance in responses that is due to linear within-person change over time. This measure gives us a sense of the importance of intrapersonal change over the course of the panel relative to the differences that exist between people assuming they never changed.

We observe similar patterns across the datasets that have produced conflicting interpretations. In all datasets we see similar rates of individual-level change over time across a broad range of measures of personal culture. But quantifying these processes also reveals that intrapersonal change plays a substantially smaller role in explaining the variance in peoples’ personal culture than between-person differences in their baselines. Of course, whether this amount of intrapersonal change is “meaningful” ultimately remains a theoretical question reflected in prior expectations concerning the question, population, and time. Nevertheless, with important exceptions, the results suggest that intrapersonal change (in some cases over more than 10 years) does not seem to play a substantial role in explaining a broad range of personal cultural differences we observe between adults.

These results help clarify and, hopefully, transcend previous debates. What is more, the measure we propose offers a tool with wider utility than merely discerning durable change that can help advance theoretical debates. To exemplify this, we turn to the role of education for cultural differences. Specifically, we use our measure to better understand what mechanisms might underlie differences observed between people with and those without college degrees. Focusing on eight questions that tab general political dispositions, our measure enables us to compare the relative importance of intrapersonal change for both groups in ways that the tournament of models cannot. We find less evidence of durable change among those with college degree. This may imply that the college experience predominantly serves as a formative period that solidifies political dispositions, rather than fostering an openness [ADAPTABILITY?] to new information and influences later in the life. [MISSING AN ENDING SENTENCE NOW.]

# Background

## Change and Stability in Personal Culture

Recent debates about whether adults undergo intrapersonal cultural change emerged in part because theories of cultural change at the aggregate level tend to implicitly invoke one of two models of individual behavior. The first, what Kiley and Vaisey [-@kiley2020] call a “Settled Dispositions Model” (SDM), assumes that peoples’ personal culture is relatively fixed by the time they are adults. While they might make temporary changes in their declarative culture in reaction to their environments, this model assumes that people return to a settled baseline over a (relatively) short period of time. This model is frequently invoked in theories of cultural change that suggest people are strongly imprinted by early socialization experiences such as the “past conditions of production” a Bourdieusian practice theory [@bourdieu1990], cohort replacement theories of aggregate change [@mannheim1952; @ryder1965], and control theories in social psychology [@smith-lovin1988; @robinson2007].

The second model summarized by Kiley and Vaisey [-@kiley2020], an “Active Updating Model” (AUM), posits that people continually update their personal culture as they move through life. This model suggests people change their personal culture as they adapt and make new meanings when encountering new social environments, discourses, and information [@gross2009; @swidler2001]. This model underlies, among others, theories of cultural diffusion [@christakis2010], attitude alignment [@dellaposta2020], and polarization [@bail2018] It is also implicit in most studies that ask whether specific experiences, changes in social roles, or political events, affect personal culture [@gelman2021; @visser2004; @slothuus2021].

There is no reason to believe that only one of these two models is “correct” at all times and in all places. A population observed over some period of time contains a mix of people who are noticeably changing and people who are not. Instead, different perspectives argue that each of these ideal-typical models is more operative at different times, for different people, and for different elements of personal culture. For example, adolescence and early adulthood is typically viewed as a “formative period” for personal cultural development and thus characterized by higher rates of active updating, while middle age and later life are potentially characterized more by settled dispositions [@alwin1991; @krosnick1989; @eaton2009]. Similarly, highly salient issues, such as views around gay rights in the 2010s; issues that see substantial elite realignments, such as views around the Vietnam conflict in the 1970s [@zaller1992]; and novel issues of public opinion, such as views around vaccines during the Covid-19 pandemic, might be characterized by active updating, while established issues of low salience are characterized by stability.

Empirically comparing which of these models better fit a broad range of questions from the General Social Survey’s rotating panels, Kiley and Vaisey [-@kiley2020] found limited evidence of durable change. While the AUM was preferred for the majority of items, the total amount of durable change detected on these items was small. A substantial minority of questions (39 percent) favored the SDM, meaning they showed no evidence of durable change. Questions with more evidence of durable change included highly salient issues like gay marriage and “public” issues like partisan identification and religious service attendance. There was also more evidence of durable change among early adults (people ages 18-30) than the rest of the population. Overall, the researchers concluded that “results ultimately suggest that real, persistent attitude change is an uncommon phenomenon among adults'' [@kiley2020: p. 500; see also @vaisey2021]. This lack of durable change is consistent with other findings that cohort replacement plays a larger role than period effects in explaining differences in personal culture [@vaisey2016].

On the other hand, the claim that change is a relatively infrequent phenomenon has been difficult to square with research identifying durable change as a result of social experiences across a number of cultural dimensions, such as morality [@brocic2021], trust [@mewes2021], and concerns about immigration [@kratz2021]. Similarly, there are longitudinal studies showing that cues from political elites can change an individual’s position on specific policy issues [@slothuus2021; @zaller1992] and that changes in their close contacts and acquaintances can change individuals' attitudes on group-related politics [@gelman2021; @dellaposta2018]. Given how often we observe people change their personal culture, it is hard to accept that adults do not change.

Drawing on these findings, Lersch challenged the SDM and AUM as a “needless dichotomy” [-@lersch2023]. Instead he proposed the “Life Course Adaption Model” (LCAM). This model draws on the life course perspective to model personal culture as a linear trajectory over the course of a panel that varies for each respondent. Lersch rectifies two shortcomings of the AUM and SDM. First, the AUM, as described by Kiley and Vaisey, posits that changes follow a Markovian process where personal culture responses at time $t$ rely solely on responses at time $t-1$. Howevre, earlier life experiences can (directly and indirectly) mold personal culture when transitioning to new social roles or into new environments, even if their initial impact is delayed. For example, childhood events might influence views on family structures later when individuals form their own families. The LCAM considers therefore considers influences from earlier than just $t-1$ on responses at time $t$.

Second, Kiley and Vaisey's analysis of the AUM and SDM is based on three-wave panel data over four years, which might not be extensive enough to properly adjudicate [COMPARE? WE ARE USING ADJUDICATING A LOT] the two models. Lersch, however, evaluates the LCAM against the AUM and SDM using panel data spanning a wider duration (from 3 to 36 years) and more waves (3 to 18), offering a better change to observe durable change.

In Lersch's tournament of models, the LCAM gains broad [EXTENSIVE?] support from data across five countries. Out of 428 questions, the LCAM was favored in 297 cases while the SDM was preferred in only 112. The rest did not yield inconclusive results, and notably, no questions favoured the AUM. He concludes that “new experiences over the life course [...] can persistently move individuals’ personal culture in novel directions” [@lersch2023: p. 243-244].

Despite differing interpretations of these scholars, their empirical results align closely. Kiley and Vaisey [-@kiley2020] discovered that most questions they investigated favored the AUM, indicating evidence of durable change among adults -- a finding largely in line with Lersch’s. However, while Lersch [-@lersch2023] found a preference for the LCAM in many cases, durable change only accounted for a minor proportion of variance explained. On average, individuals changed [CAN WE SAY SHIFTED TO INDICATE THE LINEAR DRIFT] about .07 standard deviations over 10 years, echoing Kiley's and Vaisey’s view that such changes are typically modest. Furthermore, a quarter of the items in Lersch's study favoured the SDM, suggesting that despite broader assumptions about change and more extensive data, many questions reveal no evidence of durable change. Likewise, in studies claiming evidence of durable change in personal culture, the magnitude is often minimal. For instance, Brocic and Miles [-@brocic2021] note that obtaining graduate degrees in humanities, arts, and social sciences only shift [NUDGES?] peoples’ moral relativism by about .2 standard deviations compared to non-degree holders [DELETED THE LAST SENTENCE PART HERE; FELT UNNECESSARY]. Moreover, studies of aggregate change demonstrate that even on questions where cohort effects prevail, there is always evidence that some individuals change over time [@vaisey2016]. In other words, despite different interpretations, the results of previous work are in many ways highly consonant.

## Quantification, Not Adjudication

Current debates often narrow down to ask whether people *ever* change, which is rarely theoretically interesting since the answer is almost always “yes.”  Over sufficient time and in a large enough sample, researchers will likely observe some evidence for durable changes in individuals. A lack thereof may simply due to a lack of clarity in the question asked or a low level of resolution in the possible answers, or that the question simply was not asked long enough or to enough people. Conversely, finding evidence of change tells us little about how much change is present in a population and how meaningful that change is for social differences.

A more theoretically productive approach begins with a model assuming that during an observed time period some individuals might remain stable, might change a little, or might undergo significant shifts in their personal culture. Lersch’s LCAM formally does this by modeling each individual as following their own trajectory over time. But rather than debating whether this model is superior to a model that assumes that people never change, researchers should quantify individual changes over time and compare those to the differences that exist between people at the panel's start. In other words, the focus should shift from asking, “do people change?” to asking, “how do change and initial differences collectively explain variation in personal culture?” [I CHANGED THAT QUESTION WORDING, PLEASE CHECK!]

In this approach, two metrics reflect distinct theoretical processes. First, "early imprinting," as Lersch calls it, denotes interpersonal differences at the start of the panel, capturing the accumulated experiences of people before participating in the panel that influence their answers. While commonly associated with experiences during an early formative period that result in settled disposition, this could also reflect transformative experiences [-@paul2014transformative] that happen at any stage in life, as long as they predate the panel and consistently impact subsequent responses. For instance, for those people entering the panel post-retirement, this “imprinting” might reflect this pivotal life transition. Consequently, this metric also reflects variation in individuals' social roles or statuses at the start of the panel that were important in shaping their dispositions.

The second process, termed “persistent change” or “adaption” by Lersch and "active updating" by Kiley and Vaisey, denotes durable changes in personal culture over time, captured by the amount of intrapersonal linear change during the panel. Lersch attributes these changes to social triggers such as moving into a new environment or adopting new social roles. Additionally, they might reflect the diffusion of new cultural forms across social networks, cues from political elites or otherwise culturally influential leaders, the emergence of issues in politics or culture, or large-scale social shifts.

A third metric, often termed “residual variance” or “fluctuation,” represents the remainder of variance in peoples’ responses over time. This non-durable change may arise for various reasons. Some might lack clear cultural dispositions, leading them to construct an opinion or attitude during the survey, with varying thoughts and considerations dominating their thinking in each interview [@tourangeau2000; @zaller1992; @feldman1992]. Even though the full set of considerations they could draw on remains stable over time, this could yield to such variance in responses. Additionally, this variance can also include measurement errors, such as misinterpreted questions, erroneous response selections, or responses getting coded incorrectly. While this third process thus might reflect potentially important aspects of personal culture [I FELT WE NEED TO PLAY IT DOWN A BIT], it is less central to the debate outlined here. Hence, we focus on the first two metrics.

At a theoretical level, the coexistence of two main processes -- interpersonal differences and intrapersonal change -- is undeniable, and they are linked in many ways [THOUGHT IF intrinsically linked]. Any intrapersonal change during one's life will inevitably manifest as interpersonal differences by the time people enter a panel survey. Furthermore, unless people are entirely socialized early on and never deviate from these dispositions -- an understanding challenged by evidence -- we expect observing intrapersonal change in a segment of the population when surveyed over time.

@@

## Life Course Adaption Model

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  11pt,
]{article}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={Quantifying Stability and Change in Personal Culture Using Panel Data},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin=1in]{geometry}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi

\title{Quantifying Stability and Change in Personal Culture Using Panel
Data}
\author{}
\date{\vspace{-2.5em}}

\begin{document}
\maketitle
\begin{abstract}
Recent work has produced conflicting interpretations on the question of
whether adults undergo durable changes in their personal culture over
time. This paper asserts that this divergence is less due to
researchers' findings than the approach taken -- a tournament of models
focusing on the presence or absence of change. To move beyond this, we
reanalyze the data used previously in this debate with an approach that
quantifies the processes of cultural difference. In doing so, we compare
how much cultural difference is explained by intrapersonal change and
interpersonal differences. Our results harmonize recent findings by
showing that, while most measures of personal culture show change over
time, the variance explained by intrapersonal change is typically small
compared to interpersonal differences at baseline. Quantifying these
processes offers new ways to address theoretical questions of when and
how their relative importance shifts. We examplify this with the case of
college experience.
\end{abstract}

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

A recurring focus in the sociology of culture is the question of whether
people, as they move through their lives, change their personal culture
--- the declarative and non-declarative attitudes, beliefs, values,
practices, and dispositions (Kiley and Vaisey 2020; Lersch 2023; Lizardo
2017).

This seemingly simple question underlies a number of central debates
across sociological subfields. For example, pragmatist theories of
action claim that changes in social environments cause people to adapt
their views and make new meanings (Gross 2009; Swidler 2001), while
Bourdieusian practice theories argue that the ``past conditions of
production'' leave a mark on people's personal culture that lasts
throughout their lives (Bourdieu 1990). Similarly, models of social
influence assume that people adapt their culture in the face of new
information (Vaisey and Lizardo 2010), while the emphasis on cohort
effects in models of aggregate social and cultural change requires them
to be open to change while young but become fairly resistant to it as
they age (Ryder 1965). Finally, life course theories posit slower but
still important changes over time as people advance through important
transitions in their lives (Bardi and Goodwin 2011; Elder and George
2016).

Because all of the above theoretical perspectives have some empirical
support, the current theoretical debate revolves around not whether they
exist, but when and how they each apply and the relative contribution of
each to explaining cultural differences. That being said, scholars have
struggled to reach a consensus on the relative importance of change
during adulthood. Over the span of a few years, most apparent changes in
people's responses to personal culture items appear to be transitory,
with little evidence of persistent change among adults (Kiley and Vaisey
2020; Vaisey and Kiley 2021). These results suggest change is not a
strong force in explaining personal cultural differences. When looked at
over longer time horizons, however, there is at least some evidence that
adults make persistent cultural changes, especially as they move through
distinct social roles across the life course (Lersch 2023).

These seemingly inconsistent findings are partly due to the fact that
researchers have taken a ``tournament of models'' approach in
adjudicating the different perspectives (Lersch 2023: p.~228).
Accordingly, researchers have formalized models of responses to cultural
items over time as demonstrating either ``change'' or ``no change'' to
then compare which model fits observed data better, thus functionally
asking whether we ever observe within-person change in personal culture.

We argue, however, that asking only \emph{whether} people change
obscures what are actually highly compatible empirical patterns. Whether
researchers are able to detect ``change'' in a given dataset depends not
only on sample populations, item measurement, statistical methods and
power, but crucially also on definitions of what counts as ``change.''
These methodological differences therefore limit the possibility of
convergence on the theoretical debates. Moreover, asking whether these
is evidence of change limits the kind of questions that can be asked to,
``Do we see change in this measure?'' But such inquiries are ill-suited
to address what's at the core of the theoretical debate. There is enough
evidence to assume that neither socialization at an early age nor the
influence of the contemporary environment alone can explain cultural
differences. The question that remains is under what conditions, across
which groups, and in what domains do we observe differences in the
relative importance of intrapersonal change and interpersonal
differences. This focus is necessarily comparative and requires an
approach that quantifies forms of cultural change rather than calling
victory for the one or the other perspective.

In this paper, we try to move beyond previous debates by moving away
from the ``tournament of models'' approach toward a quantification of
the processes that lead to differences in personal culture. Drawing on
the seven panel datasets analyzed in these debates, we use established
methods to quantify the amount of variance in responses that is due to
between-people differences at a single point during the survey versus
the amount of variance in responses that is due to linear within-person
change over time. This measure gives us a sense of the importance of
intrapersonal change over the course of the panel relative to the
differences that exist between people assuming they never changed.

We observe similar patterns across the datasets that have produced
conflicting interpretations. In all datasets we see similar rates of
individual-level change over time across a broad range of measures of
personal culture. But quantifying these processes also reveals that
intrapersonal change plays a substantially smaller role in explaining
the variance in peoples' personal culture than between-person
differences in their baselines. Of course, whether this amount of
intrapersonal change is ``meaningful'' ultimately remains a theoretical
question reflected in prior expectations concerning the question,
population, and time. Nevertheless, with important exceptions, the
results suggest that intrapersonal change (in some cases over more than
10 years) does not seem to play a substantial role in explaining a broad
range of personal cultural differences we observe between adults.

These results help clarify and, hopefully, transcend previous debates.
What is more, the measure we propose offers a tool with wider utility
than merely discerning durable change that can help advance theoretical
debates. To exemplify this, we turn to the role of education for
cultural differences. Specifically, we use our measure to better
understand what mechanisms might underlie differences observed between
people with and those without college degrees. Focusing on eight
questions that tab general political dispositions, our measure enables
us to compare the relative importance of intrapersonal change for both
groups in ways that the tournament of models cannot. We find less
evidence of durable change among those with college degree. This may
imply that the college experience predominantly serves as a formative
period that solidifies political dispositions, rather than fostering an
openness {[}ADAPTABILITY?{]} to new information and influences later in
the life. {[}MISSING AN ENDING SENTENCE NOW.{]}

\hypertarget{background}{%
\section{Background}\label{background}}

\hypertarget{change-and-stability-in-personal-culture}{%
\subsection{Change and Stability in Personal
Culture}\label{change-and-stability-in-personal-culture}}

Recent debates about whether adults undergo intrapersonal cultural
change emerged in part because theories of cultural change at the
aggregate level tend to implicitly invoke one of two models of
individual behavior. The first, what Kiley and Vaisey (2020) call a
``Settled Dispositions Model'' (SDM), assumes that peoples' personal
culture is relatively fixed by the time they are adults. While they
might make temporary changes in their declarative culture in reaction to
their environments, this model assumes that people return to a settled
baseline over a (relatively) short period of time. This model is
frequently invoked in theories of cultural change that suggest people
are strongly imprinted by early socialization experiences such as the
``past conditions of production'' a Bourdieusian practice theory
(Bourdieu 1990), cohort replacement theories of aggregate change
(Mannheim 1952; Ryder 1965), and control theories in social psychology
(Robinson 2007; Smith-Lovin and Heise 1988).

The second model summarized by Kiley and Vaisey (2020), an ``Active
Updating Model'' (AUM), posits that people continually update their
personal culture as they move through life. This model suggests people
change their personal culture as they adapt and make new meanings when
encountering new social environments, discourses, and information (Gross
2009; Swidler 2001). This model underlies, among others, theories of
cultural diffusion (Christakis and Fowler 2010), attitude alignment
(DellaPosta 2020), and polarization (Bail et al. 2018) It is also
implicit in most studies that ask whether specific experiences, changes
in social roles, or political events, affect personal culture (Gelman
and Margalit 2021; Slothuus and Bisgaard 2021; Visser and Mirabile
2004).

There is no reason to believe that only one of these two models is
``correct'' at all times and in all places. A population observed over
some period of time contains a mix of people who are noticeably changing
and people who are not. Instead, different perspectives argue that each
of these ideal-typical models is more operative at different times, for
different people, and for different elements of personal culture. For
example, adolescence and early adulthood is typically viewed as a
``formative period'' for personal cultural development and thus
characterized by higher rates of active updating, while middle age and
later life are potentially characterized more by settled dispositions
(Alwin and Krosnick 1991; Eaton et al. 2009; Krosnick and Alwin 1989).
Similarly, highly salient issues, such as views around gay rights in the
2010s; issues that see substantial elite realignments, such as views
around the Vietnam conflict in the 1970s (Zaller 1992); and novel issues
of public opinion, such as views around vaccines during the Covid-19
pandemic, might be characterized by active updating, while established
issues of low salience are characterized by stability.

Empirically comparing which of these models better fit a broad range of
questions from the General Social Survey's rotating panels, Kiley and
Vaisey (2020) found limited evidence of durable change. While the AUM
was preferred for the majority of items, the total amount of durable
change detected on these items was small. A substantial minority of
questions (39 percent) favored the SDM, meaning they showed no evidence
of durable change. Questions with more evidence of durable change
included highly salient issues like gay marriage and ``public'' issues
like partisan identification and religious service attendance. There was
also more evidence of durable change among early adults (people ages
18-30) than the rest of the population. Overall, the researchers
concluded that ``results ultimately suggest that real, persistent
attitude change is an uncommon phenomenon among adults'\,' (Kiley and
Vaisey 2020: p.~500; see also Vaisey and Kiley 2021). This lack of
durable change is consistent with other findings that cohort replacement
plays a larger role than period effects in explaining differences in
personal culture (Vaisey and Lizardo 2016).

On the other hand, the claim that change is a relatively infrequent
phenomenon has been difficult to square with research identifying
durable change as a result of social experiences across a number of
cultural dimensions, such as morality (Broćić and Miles 2021), trust
(Mewes et al. 2021), and concerns about immigration (Kratz 2021).
Similarly, there are longitudinal studies showing that cues from
political elites can change an individual's position on specific policy
issues (Slothuus and Bisgaard 2021; Zaller 1992) and that changes in
their close contacts and acquaintances can change individuals' attitudes
on group-related politics (DellaPosta 2018; Gelman and Margalit 2021).
Given how often we observe people change their personal culture, it is
hard to accept that adults do not change.

Drawing on these findings, Lersch challenged the SDM and AUM as a
``needless dichotomy'' (2023). Instead he proposed the ``Life Course
Adaption Model'' (LCAM). This model draws on the life course perspective
to model personal culture as a linear trajectory over the course of a
panel that varies for each respondent. Lersch rectifies two shortcomings
of the AUM and SDM. First, the AUM, as described by Kiley and Vaisey,
posits that changes follow a Markovian process where personal culture
responses at time \(t\) rely solely on responses at time \(t-1\).
Howevre, earlier life experiences can (directly and indirectly) mold
personal culture when transitioning to new social roles or into new
environments, even if their initial impact is delayed. For example,
childhood events might influence views on family structures later when
individuals form their own families. The LCAM considers therefore
considers influences from earlier than just \(t-1\) on responses at time
\(t\).

Second, Kiley and Vaisey's analysis of the AUM and SDM is based on
three-wave panel data over four years, which might not be extensive
enough to properly adjudicate {[}COMPARE? WE ARE USING ADJUDICATING A
LOT{]} the two models. Lersch, however, evaluates the LCAM against the
AUM and SDM using panel data spanning a wider duration (from 3 to 36
years) and more waves (3 to 18), offering a better change to observe
durable change.

In Lersch's tournament of models, the LCAM gains broad {[}EXTENSIVE?{]}
support from data across five countries. Out of 428 questions, the LCAM
was favored in 297 cases while the SDM was preferred in only 112. The
rest did not yield inconclusive results, and notably, no questions
favoured the AUM. He concludes that ``new experiences over the life
course {[}\ldots{]} can persistently move individuals' personal culture
in novel directions'' (Lersch 2023: p.~243-244).

Despite differing interpretations of these scholars, their empirical
results align closely. Kiley and Vaisey (2020) discovered that most
questions they investigated favored the AUM, indicating evidence of
durable change among adults -- a finding largely in line with Lersch's.
However, while Lersch (2023) found a preference for the LCAM in many
cases, durable change only accounted for a minor proportion of variance
explained. On average, individuals changed {[}CAN WE SAY SHIFTED TO
INDICATE THE LINEAR DRIFT{]} about .07 standard deviations over 10
years, echoing Kiley's and Vaisey's view that such changes are typically
modest. Furthermore, a quarter of the items in Lersch's study favoured
the SDM, suggesting that despite broader assumptions about change and
more extensive data, many questions reveal no evidence of durable
change. Likewise, in studies claiming evidence of durable change in
personal culture, the magnitude is often minimal. For instance, Brocic
and Miles (2021) note that obtaining graduate degrees in humanities,
arts, and social sciences only shift {[}NUDGES?{]} peoples' moral
relativism by about .2 standard deviations compared to non-degree
holders {[}DELETED THE LAST SENTENCE PART HERE; FELT UNNECESSARY{]}.
Moreover, studies of aggregate change demonstrate that even on questions
where cohort effects prevail, there is always evidence that some
individuals change over time (Vaisey and Lizardo 2016). In other words,
despite different interpretations, the results of previous work are in
many ways highly consonant.

\hypertarget{quantification-not-adjudication}{%
\subsection{Quantification, Not
Adjudication}\label{quantification-not-adjudication}}

Current debates often narrow down to ask whether people \emph{ever}
change, which is rarely theoretically interesting since the answer is
almost always ``yes.'' Over sufficient time and in a large enough
sample, researchers will likely observe some evidence for durable
changes in individuals. A lack thereof may simply due to a lack of
clarity in the question asked or a low level of resolution in the
possible answers, or that the question simply was not asked long enough
or to enough people. Conversely, finding evidence of change tells us
little about how much change is present in a population and how
meaningful that change is for social differences.

A more theoretically productive approach begins with a model assuming
that during an observed time period some individuals might remain
stable, might change a little, or might undergo significant shifts in
their personal culture. Lersch's LCAM formally does this by modeling
each individual as following their own trajectory over time. But rather
than debating whether this model is superior to a model that assumes
that people never change, researchers should quantify individual changes
over time and compare those to the differences that exist between people
at the panel's start. In other words, the focus should shift from
asking, ``do people change?'' to asking, ``how do change and initial
differences collectively explain variation in personal culture?'' {[}I
CHANGED THAT QUESTION WORDING, PLEASE CHECK!{]}

In this approach, two metrics reflect distinct theoretical processes.
First, ``early imprinting,'' as Lersch calls it, denotes interpersonal
differences at the start of the panel, capturing the accumulated
experiences of people before participating in the panel that influence
their answers. While commonly associated with experiences during an
early formative period that result in settled disposition, this could
also reflect transformative experiences (2014) that happen at any stage
in life, as long as they predate the panel and consistently impact
subsequent responses. For instance, for those people entering the panel
post-retirement, this ``imprinting'' might reflect this pivotal life
transition. Consequently, this metric also reflects variation in
individuals' social roles or statuses at the start of the panel that
were important in shaping their dispositions.

The second process, termed ``persistent change'' or ``adaption'' by
Lersch and ``active updating'' by Kiley and Vaisey, denotes durable
changes in personal culture over time, captured by the amount of
intrapersonal linear change during the panel. Lersch attributes these
changes to social triggers such as moving into a new environment or
adopting new social roles. Additionally, they might reflect the
diffusion of new cultural forms across social networks, cues from
political elites or otherwise culturally influential leaders, the
emergence of issues in politics or culture, or large-scale social
shifts.

A third metric, often termed ``residual variance'' or ``fluctuation,''
represents the remainder of variance in peoples' responses over time.
This non-durable change may arise for various reasons. Some might lack
clear cultural dispositions, leading them to construct an opinion or
attitude during the survey, with varying thoughts and considerations
dominating their thinking in each interview (Feldman and Zaller 1992;
Tourangeau, Rips, and Rasinski 2000; Zaller 1992). Even though the full
set of considerations they could draw on remains stable over time, this
could yield to such variance in responses. Additionally, this variance
can also include measurement errors, such as misinterpreted questions,
erroneous response selections, or responses getting coded incorrectly.
While this third process thus might reflect potentially important
aspects of personal culture {[}I FELT WE NEED TO PLAY IT DOWN A BIT{]},
it is less central to the debate outlined here. Hence, we focus on the
first two metrics.

At a theoretical level, the coexistence of two main processes --
interpersonal differences and intrapersonal change -- is undeniable, and
they are linked in many ways {[}THOUGHT IF intrinsically linked{]}. Any
intrapersonal change during one's life will inevitably manifest as
interpersonal differences by the time people enter a panel survey.
Furthermore, unless people are entirely socialized early on and never
deviate from these dispositions -- an understanding challenged by
evidence -- we expect observing intrapersonal change in a segment of the
population when surveyed over time.

@@

\hypertarget{life-course-adaption-model}{%
\subsection{Life Course Adaption
Model}\label{life-course-adaption-model}}
